---
title: "Portfolio Project"
author: "Tyler Busby"
data: "4-10-2025"
format: html
---

```{r, include=FALSE}
#| label: "load-packages"

library(tidyverse)
library(ggimage)
library(ggraph)
library(igraph)
library(ggridges)
```

# Data Description

The data I will be using was gathered from user observations on iNaturalist, using the following website: https://www.inaturalist.org/observations/export.

The data focuses on nudibranch observations in Monterey Bay Marine Sanctuary from January 1st to December 31st, 2024. It covers all user-submitted observations of animals in the order Nudibranchia that are considered "research grade", meaning the nudibranch ID was confirmed by at least one other user. The data also only includes observations marked as open, meaning they have an associated location.

The dataset includes a number of variables, including observation id variables: id, uuid, user_id, user_name, and url. Time and date variables for the observation, including observed_on, time_observed_at, and time_zone. There are also location variables: place_guess, latitude, longitude, and positional_accuracy. The main variables I will focus on have to do with the observation's taxonomy including it's common_name, scientific_name, taxon_genus_name, taxon_family_name, and more for every taxon delineation from suborder to form. The number of agreements and disagreements along with other information included by the observer, such as tag_list and description.

Most of the data is character data, with the exception of observed_on, which is a date, and number_identification_agreements, number_identification_disagreements, latitude, longitude, and positional_accuracy, which are numerical variables.

The main questions I hope to answer with my data are:\
- What were the most commonly-observed nudibranch species in Monterey Bay in 2024?\
- How biodiverse were nudibranch populations in Monterey Bay in 2024?\
- How do nudibranch populations change in Monterey Bay over time?

# Data Cleaning

In order to create my first visualization, I needed to first create a new dataset from the original nudibranch dataset. Because a few species of nudibranch do not have a common name, I needed to use the coalesce() function to replace any common names that were marked as NA with the nudibranch's scientific name. This allows us to separate these nudibranchs into distinct species instead of accidentally lumping them all together. I then summarized the total number of observations for each species of nudibranch, giving us the total observation count for each distinct species. Finally, using a piece of code modeled after a tutorial from Dean Chereden on YouTube, I added photos to each of the ten most common species of nudibranch to be used in the first visualization.

To create my second visualization, in order to create a dendrogram of all the species of nudibranchs, I needed to create a dataset with each of the individual nodes and their origins. To accomplish this, I created 5 datasets where common name, genus, family, superfamily, and suborder each served as nodes and combined them into one dataset using rbind(). I then used this dataset to form a graph object to be used in my dendrogram.

For my final visualization, much like my first, I used the the coalesce() function to replace any common names that were marked as NA. I then used a mutate() function to extract the month from the date variable.

```{r, include=FALSE}
#| label: "load-data"

nudibranch <- read_csv("data-raw/nudibranch-obs.csv")
```

```{r, include=FALSE}
#| label: "creating-totals-dataset"

tot_nudibranch <- nudibranch |> #creates new tot_nudibranch dataset
  mutate(common_name = coalesce(common_name, scientific_name)) |> #replaces nas in common name variable with scientific name
  group_by(common_name) |> #group by common_name
  summarise("total_obs" = length(common_name)) #summarizes total_obs as the number of each common name observation

tot_nudibranch$image_file <- paste0("data-raw/nudibranch-photos/", 
                                     tot_nudibranch$common_name, ".png") #creates image address variable for each common name (see below)

write_csv(tot_nudibranch, file = 'data-clean/tot_nudibranchs.csv') #save new totals dataset as updated csv file
```

# Data Visualization 1

The first visualization is a bar chart of the 10 species with the highest number of observations. The visualization clearly shows readers which species have the highest and lowest number of observations by placing the barsin order of size. Not to mention, drawings of each of the nudibranchs were included in order help the readers connect with the data further by allowing them to see what each of these species look like. As shown in the graph, the most common species of nudibranch in the Monterey Bay is the opalescent nudibranch, followed by the spotted dorid, Hopkin's rose, Monterey dorid, San Diego dorid, clown dorid, Hilton's aeolid, noble dorid, Heath's dorid, and Cockerell's dorid.

```{r, echo=FALSE}
#| label: "first-visualization-rough-draft"

tot_nudibranch |>
  arrange(-total_obs) |> #arrange by descending number of observations of each species
  head(10) |> #select only top 10 observations
  ggplot(aes(x = total_obs, 
               y = fct_reorder(common_name, total_obs))) + #create plot with given aesthetics
  geom_col(aes(fill = total_obs), 
           color = "black") + #create bar plot with color based on observation number
  guides(fill = "none") + #remove legend
  labs(x = "Total Number of iNaturalist Observations in 2024",
       title = "10 Most Common Nudibranch Species in Monterey Bay", #create descriptive labels
       alt = "Bar plot with nudibranch species on the y axis and total number of observations on the x-axis. X-axis ranges from 0-1200, with the nudibranch with the highest number of observations being the opalescent nudibranch.") + #add alt text
  scale_x_continuous(expand = c(0,0), 
                     limits = c(0,1250),
                     breaks = c(0, 200, 400, 600, 800, 1000, 1200)) + #resize x axis scale
  geom_image(aes(image = image_file),
             size = 0.16) + #insert image on each bar in the bar plot (see below)
  theme_classic() + #set theme to classic
  theme(axis.title.y = element_blank(), #remove y axis label
        title = element_text(family = "serif"),
        axis.text.x = element_text(family = "serif"),
        axis.text.y = element_text(family = "serif"), #change font for all texts
        panel.background = element_rect(fill = '#D9E3F1'),
        plot.background = element_rect(fill = '#D9E3F1')) + #change bg color to match website
  scale_fill_distiller(direction = 1) #change fill color to brewer blues
```

Barplot of nudibranch observations

# Data Visualization 2

The second visualization shows a dendrogram off all of the species of nudibranch found in the Monterey Bay area. It allows readers to observe the biodiversity of nudibranch species, genera, families, superfamilies, and suborders by showing them as nodes or points of divergence. The visualization shows readers each of the 75 species of nudibranch observed in the Monterey Bay area in 2024, along with the 47 unique genera, 31 unique families, 11 unique superfamilies, and 2 unique suborders of nudibranch in Monterey Bay along with how each one relates to each other. This specific dendrogram's code was modeled off of a tutorial on the R Graph Gallery.

```{r, include=FALSE}
#| label: "creating-nudi_edges-dataset"

nudi1 <- nudibranch |> #creates nudi1 dataframe using mollusca as the origin and suborder as first group
  select(iconic_taxon_name, taxon_suborder_name) |> #picking only relevant variables
  unique() |> #selecting only each unique occurance
  rename("from" = iconic_taxon_name, "to" = taxon_suborder_name) #renaming variables for later cbind

nudi2 <- nudibranch |>
  select(taxon_suborder_name, taxon_superfamily_name) |>
  unique() |>
  rename("from" = taxon_suborder_name, "to" = taxon_superfamily_name) #repeating above for suborder and superfamily

nudi3 <- nudibranch |>
  select(taxon_superfamily_name, taxon_family_name) |>
  unique() |>
  rename("from" = taxon_superfamily_name, "to" = taxon_family_name) #repeating above for superfamily and family

nudi4 <- nudibranch |>
  select(taxon_family_name, taxon_genus_name) |>
  unique() |>
  rename("from" = taxon_family_name, "to" = taxon_genus_name) #repeating above for family and genus

nudi5 <- nudibranch |>
  mutate(common_name = coalesce(common_name, scientific_name)) |> #adding scientific to NA common names
  select(taxon_genus_name, common_name) |>
  unique() |>
  rename("from" = taxon_genus_name, "to" = common_name) #repeating above for genus and common name (species)

nudi_edges <- rbind(nudi1, nudi2, nudi3, nudi4, nudi5) #combining all 5 datasets into one edge dataset
```

```{r, echo=FALSE}
nudi_dendrogram <- graph_from_data_frame(nudi_edges) #creating a graph object to be used in ggraph

ggraph(nudi_dendrogram, layout = 'dendrogram', circular = FALSE) + #creating dendrogram from nudi_dendrogram
  geom_edge_diagonal() +
  geom_node_text(aes(label=name, filter=leaf), 
                 angle=90 , 
                 hjust=1, 
                 nudge_y = -0.1, 
                 nudge_x = -0.15,
                 size = 2) + #creates text object for each common name and aligns it with node point
  geom_node_point(aes(filter=leaf) , alpha=0.6) + #creates node point at each end
  ylim(-2.15, NA) + #expands y-axis vertically
  theme_void() + #sets theme to void
  theme(panel.background = element_rect(fill = '#D9E3F1'),
        plot.background = element_rect(fill = '#D9E3F1')) + #change bg color to match website)
  labs(alt = "Dendrogram showing each of the 75 species of nudibranch reported in Monterey Bay in 2024. Each species has a node which branch from a central origin, with branches appearing at the genus, family, subfamily, and suborder level. The species are split into 47 unique genera, 31 unique families, 11 unique superfamilies, and 2 unique suborders.") #add alt text)
```

Dendrogram of nudibranchs in the Monterey Bay

# Data Visualization 3

The final visualization is a ridgeline plot of nudibranch observations over each of the 12 months by species. This helps readers notice the seasonal trends of nudibranch populations, generally reaching their peak around May, June, or July, and reaching a lull from August to April (with some exceptions, of course). Readers may also identify a potential peak around October to December for some species in particular.

```{r, include=FALSE}
monthly_nudibranch <- nudibranch |> #creates new monthly_nudibranch dataset
  mutate(common_name = coalesce(common_name, scientific_name), #replaces nas in common name variable with scientific name
         month = month(observed_on, label = FALSE)) #summarizes total_obs as the number of each common name observation
```

```{r, echo=FALSE}
ggplot(monthly_nudibranch, aes(x = month, y = common_name, fill = common_name)) +
  geom_density_ridges(trim = TRUE, 
                      scale = 5) + 
  theme_ridges() + #create ridge plot with x of month and y of common_name
  theme(axis.text.y = element_blank(),
        axis.title.y = element_text(size = 1),
        axis.text.x = element_text(size = 6,
                                   family = "serif"),
        axis.title.x = element_text(hjust = 0.5,
                                    family = "serif"),
        title = element_text(size = 10, 
                             family = "serif"),  #change font and size for all texts
        legend.position = "none", #remove legend
        panel.background = element_rect(fill = '#D9E3F1'),
        plot.background = element_rect(fill = '#D9E3F1')) + #change bg color to match website)
  scale_x_continuous(expand = c(0,0),
                     limits = c(1, 12),
                     breaks = seq_along(month.name), 
                     labels = month.name) + #change scale of x axis
  labs(y = "",
       x = "Month",
       title = "Nudibranch Populations in the Monterey Bay by Month", #add descriptive labels 
       alt = "Ridgeline plot showing the each of the 75 species of nudibranch reported in Monterey Bay in 2024 on the y-axis and the month on the x-axis. The concentration of reports during each of the months for each species is shown, displaying an increase in reports during May to July and October to December. Other months, number of reports appear to be at a lull for most species with some exceptions.") + #add alt text)
  scale_fill_viridis_d(option = "mako")
```

Ridgeline plot of nudibranch populations by month

# Citations

Dean Chereden. Add IMAGES to Charts in RStudio \[video\]. Youtube. 2023 Apr 27, 8:16 minutes. \[accessed 2025 May 1\]. https://www.youtube.com/watch?app=desktop&v=Mt-lVqqZCVw.

Yan Holtz. c2016-2018. Dendrogram \[Internet\]. The R Graph Gallery. \[updated 2025; cited 2025 May 15\]. Available from: <https://r-graph-gallery.com/dendrogram.html.>
